Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-01-07 supplement.tex
some clean up

$F_{ST}$ represents the
proportion of genetic variance in a multi-population sample that can be
attributed to drift between populations in that sample (e.g., \cite{Bhatia2013-sq}).
$F_{ST}$ depends on population divergence times, gene flow,
and population sizes. For present-day sampled populations,
predicted $F_{ST}$ can be compared to data (Fig.~\ref{fig:supp-FST-panels}), while past estimates represent model predictions about the differentiation between past populations. We
computed $F_{ST}$ as a ratio of expected (or average) variances, using observed
pairwise diversity within and between populations as \sgcomment{WHich estimator is that? If this is the one recommended in Bhatia, cite?}
\begin{equation}
    F_{ST}(i, j) = \frac{\E[2H_{i,j} - H_i - H_j]}{\E[2H_{i,j} + H_i + H_j]}.
    \label{eq:FST}
\end{equation}
\sgcomment{where $H_{ij}$ is the average number of differences between one haploid copy of the genome from population $i$ and one from population $j$.}
In all inferred models, $F_{ST}$s between Neanderthals and the ancestors of
modern humans were much larger than between any pair of modern human
populations or their ancestors (stems). $F_{ST}$ between modern human
populations over the last $100ka$ remained relatively low, at
$\sim0.1$ or less. In the continuous-migration model, the two stem populations
also were predicted to have low $F_{ST}$ within this range, as migration rates
were inferred to be high enough to maintain genetic similarity. In both the
merger-without-stem-migration and the merger-with-stem-migration, one of the
stem populations was inferred to have a short, severe bottleneck, which led to
increased $F_{ST}$ between stem populations during that time. In the merger with stem migration, $F_{ST}$ reached a maximum value of $0.3$ between stems, intermediate between human-Neanderthal divergence ($0.54$ \sgcomment{check}) and the maximum $F_{ST}$ observed between contemporary human populations ($0.07$.)

Whereas the high $F_{ST}$ value between Neanderthal and humans reflects both elevated genetic differences between the genomes of modern humans and neanderthal and reduced diversity within the Neanderthal population (\sgcomment{Figure X}), the increased $F_{ST}$ between the stems is primarily due to an inferred bottleneck reducing stem diversity, while genetic differentiation between individuals from the two stems is within the range of genetic differentiation between (and even within) human populations. In other words, our best fit model describes two stems with one stems showing reduced diversity, but individuals across stems showing no more differentiation than two modern population.  \sgcomment{Previous paragraph could use rereading.}

\subsection{$f_4$ statistics between pairs of contemporary and pairs of ancient populations}
\label{sec:f4}

@@

\begin{table}[ht]
\caption{
    \label{tab:supp-continuous-migration}
    \textbf{Best-fit parameters from the Continuous-Migration model.}
    Inferred values are scaled to physical units assuming a generation time of
    29 years. This model gave a log-likelihood of -116,000. \aprcomment{See
    figure XX with all parameters labeled.} 
}
\centering
\begin{tabular}[t]{rP{8cm}cc}
    \toprule
    Parameter & Description & Value & Std. err.\\
    \midrule
    $N_e$ & Ancestral effective population size & 7270 & 1777 \\
    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal and Nama splits & 8256 & 1612 \\
    $N_{stem2}$ & Size of stem 2 lineage & 13547 & 2488 \\
    $N_{Nama_0}$ & Initial Nama size & 11939 & 2989 \\
    $N_{Nama_F}$ & Final Nama size & 221 & 54 \\
    $N_{MSL_0}$ & Initial Mende size & 9738 & 2479 \\
    $N_{MSL_F}$ & Final Mende size & 28150 & 6628 \\
    $N_{EA}$ & Size of East African branch & 7489 & 1841 \\
    $N_{Gumuz_F}$ & Final Gumuz size & 3728 & 915 \\
    $N_{EP}$ & East African agriculturist size & 13072 & 3246 \\
    $N_{GBR_0}$ & Initial British size & 959 & 231 \\
    $N_{GBR_F}$ & Final British size & 11822 & 2889 \\
    $N_{Neand}$ & Neanderthal size & 2670 & 591 \\
    $T_{stems}$ & Stem split time (years) & 1163072 & 390803 \\
    $T_{Nama}$ & Nama split time (years) & 134745 & 17775 \\
    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.98\times10^{-5}$ & $0.366\times10^{-5}$ \\
    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.08\times10^{-5}$ & $1.02\times10^{-5}$ \\
    $m_{MSL-EA}$ & Mende--East Africa migration rate & $21.4\times10^{-5}$ & $5.32\times10^{-5}$ \\
    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.17\times10^{-5}$ & $1.02\times10^{-5}$ \\
    $m_{EA-EA}$ & Intra-East Africa migration rate & $33.6\times10^{-5}$ & $8.35\times10^{-5}$ \\
    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12 ka ($1-f$ from Gumuz) & 0.642 & 0.0037 \\
    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2 ka & 0.255 & 0.0043 \\
    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.156 & 0.0021 \\
    $m_{stems}$ & Stem 1--stem 2 migration rate & $6.43\times10^{-5}$ & $1.05\times10^{-5}$ \\
    $m_{stem2-Nama}$ & Stem 2--Nama migration rate & $5.82\times10^{-5}$ & $1.60\times10^{-5}$ \\
    $m_{stem2-MSL}$ & Stem 2--Mende migration rate & $16.4\times10^{-5}$ & $4.19\times10^{-5}$ \\
    $m_{stem2-EA}$ & Stem 2--East Africa migration rate & $3.10\times10^{-5}$ & $0.901\times10^{-5}$ \\
    \bottomrule
\end{tabular}
\end{table}

>>>>

$F_{ST}$ represents the proportion of genetic variance in a multi-population
sample that can be attributed to drift between populations in that sample
(e.g., \cite{Bhatia2013-sq}).  $F_{ST}$ depends on population divergence times,
gene flow, and population sizes. For present-day sampled populations, predicted
$F_{ST}$ can be compared to data (Fig.~\ref{fig:supp-FST-panels}), while past
estimates represent model predictions about the differentiation between past
populations. We computed $F_{ST}$ as a ratio of expected (or average)
variances, using observed pairwise diversity within and between populations as
in \citet{Peter2016-ys}:
\begin{equation}
    F_{ST}(i, j) = \frac{\E[2H_{i,j} - H_i - H_j]}{\E[2H_{i,j} + H_i + H_j]},
    \label{eq:FST}
\end{equation}

where $H_{ij}$ is the average number of differences between one haploid copy of
the genome from population $i$ and one from population $j$, and $H_i$ and $H_j$
are the average number of differences for two haploid copies from the same
population. In all inferred models, $F_{ST}$s between Neanderthals and the
ancestors of modern humans were much larger than between any pair of modern
human populations or their ancestors (stems). $F_{ST}$ between modern human
populations over the last $100ka$ remained relatively low, at $\sim0.1$ or
less. In the continuous-migration model, the two stem populations also were
predicted to have low $F_{ST}$ within this range, as migration rates were
inferred to be high enough to maintain genetic similarity. In both the
merger-without-stem-migration and the merger-with-stem-migration, one of the
stem populations was inferred to have a short, severe bottleneck, which led to
increased $F_{ST}$ between stem populations during that time. In the merger
with stem migration, $F_{ST}$ reached a maximum value of $0.3$ between stems,
intermediate between human-Neanderthal divergence ($0.54$ \sgcomment{check})
and the maximum $F_{ST}$ observed between contemporary human populations
($0.07$.)

Whereas the high $F_{ST}$ value between Neanderthal and humans reflects both
elevated genetic differences between the genomes of modern humans and
neanderthal and reduced diversity within the Neanderthal population
(Fig.~\ref{fig:supp-FST-panels}), the increased $F_{ST}$ between the stems is
primarily due to an inferred bottleneck reducing stem diversity, while genetic
differentiation between individuals from the two stems is within the range of
genetic differentiation between (and even within) human populations. In other
words, our best fit model describes two stems with one stems showing reduced
diversity, but individuals across stems showing no more differentiation than
two modern population. \sgcomment{Previous paragraph could use rereading.}

\subsection{\bm{$f_4$} statistics between pairs of contemporary
and pairs of ancient populations}
\label{sec:f4}

@@

\begin{table}[ht]
\caption{
    \label{tab:supp-continuous-migration}
    \textbf{Best-fit parameters from the Continuous-Migration model.}
    Inferred values are scaled to physical units assuming a generation time of
    29 years. This model gave a log-likelihood of -115,500. \aprcomment{See
    figure XX with all parameters labeled.} 
}
\centering
\begin{tabular}[t]{rP{8cm}cc}
    \toprule
    Parameter & Description & Value & Std. err.\\
    \midrule
    $N_e$ & Ancestral effective population size & 7270 & 1777 \\
    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal and Nama splits & 8256 & 1612 \\
    $N_{stem2}$ & Size of stem 2 lineage & 13547 & 2488 \\
    $N_{Nama_0}$ & Initial Nama size & 11939 & 2989 \\
    $N_{Nama_F}$ & Final Nama size & 221 & 54 \\
    $N_{MSL_0}$ & Initial Mende size & 9738 & 2479 \\
    $N_{MSL_F}$ & Final Mende size & 28150 & 6628 \\
    $N_{EA}$ & Size of East African branch & 7489 & 1841 \\
    $N_{Gumuz_F}$ & Final Gumuz size & 3728 & 915 \\
    $N_{EP}$ & East African agriculturist size & 13072 & 3246 \\
    $N_{GBR_0}$ & Initial British size & 959 & 231 \\
    $N_{GBR_F}$ & Final British size & 11822 & 2889 \\
    $N_{Neand}$ & Neanderthal size & 2670 & 591 \\
    $T_{stems}$ & Stem split time (years) & 1163072 & 390803 \\
    $T_{Nama}$ & Nama split time (years) & 134745 & 17775 \\
    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.98\times10^{-5}$ & $0.366\times10^{-5}$ \\
    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.08\times10^{-5}$ & $1.02\times10^{-5}$ \\
    $m_{MSL-EA}$ & Mende--East Africa migration rate & $21.4\times10^{-5}$ & $5.32\times10^{-5}$ \\
    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.17\times10^{-5}$ & $1.02\times10^{-5}$ \\
    $m_{EA-EA}$ & Intra-East Africa migration rate & $33.6\times10^{-5}$ & $8.35\times10^{-5}$ \\
    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12 ka ($1-f$ from Gumuz) & 0.642 & 0.0037 \\
    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2 ka & 0.255 & 0.0043 \\
    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.156 & 0.0021 \\
    $m_{stems}$ & Stem 1--stem 2 migration rate & $6.43\times10^{-5}$ & $1.05\times10^{-5}$ \\
    $m_{stem2-Nama}$ & Stem 2--Nama migration rate & $5.82\times10^{-5}$ & $1.60\times10^{-5}$ \\
    $m_{stem2-MSL}$ & Stem 2--Mende migration rate & $16.4\times10^{-5}$ & $4.19\times10^{-5}$ \\
    $m_{stem2-EA}$ & Stem 2--East Africa migration rate & $3.10\times10^{-5}$ & $0.901\times10^{-5}$ \\
    \bottomrule
\end{tabular}
\end{table}
